White Dwarf Mergers on Adaptive Meshes\\ I. Methodology and Code Verification

### 2015-09-17 papers/wdmerger_I/paper.tex
Add an appendix demonstrating that the time-centered energy update conserves total energy in simulations with self-gravity.

We use the unsplit piecewise-parabolic method (PPM) solver in \castro\
to advance the hydrodynamics system in time \citep{ppmunsplit}.  A
number of changes were made to the solver, which are detailed in the
Appendix.  These changes bring the algorithm more in line with that of
\cite{ppm}. \castro\ as originally released featured a slightly modified
version of the higher resolution limiters of
\cite{colella_sekora:2008}, which can be accessed in the code using
\texttt{castro.ppm\_type = 2}. The advantage of this limiter is that
it preserves physical extrema rather than clipping them off as in the
original approach of \cite{ppm}. Despite the advantages of this limiter 
we have found it to be unsatisfactory for our problem. There are many regions in our
problem with large density gradients (such as the interface between
the star's atmosphere and the ambient gas outside of it) and in these
regions the algorithm can yield negative densities. This often results
from the limiters interpreting these gradients as being true
minima. As a result, we use the original limiter, which is strictly
monotonicity preserving in the parabolic profiles it generates; this
is activated with \texttt{castro.ppm\_type = 1}.

@@

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
the appendix. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
We believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

@@

Finally, then, we write the update in a zone for the total energy that conserves $(\rho E_{\text{tot}})$ as:
\begin{equation}
  \Delta (\rho E) = -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi_{f+1/2} - \Phi_{f-1/2}),\label{eq:grav_energy_conservation_update}
\end{equation}
where the sum is over the cell faces with indices $f$ and the indices $f+1/2$ and $f-1/2$ refer to 
the zone centers immediately to the left and right in the direction perpendicular to the face.
As long as we record the hydrodynamical fluxes through the zone faces after coming out of the hydrodynamics step, 
this algorithm is able to conserve the total energy completely (except for any energy loss or gain through 
physical domain boundaries). In order for the method to be second-order accurate in time, 
we need to use a time-centered $\Phi$ (which can be computed by averaging the time-level $n$ and $n+1$ potentials;
we already have the latter because \castro\ re-computes the potential at the new time after the hydrodynamics step,
and we can apply this energy at the end of the timestep). Note that of course the hydrodynamical
flux is already second-order accurate in time. We observe also that in practice we will not obtain 
conservation of energy to machine precision even in the absence of open domain boundaries. The 
method itself is conservative if it is time-centered and correctly evaluates the energy change 
on cell faces \citep{marcello:2012}. However, in practice there is a non-zero numerical tolerance associated 
with the Poisson gravity solver (in our case, the multigrid method) that results in a non-zero error 
in the calculation of the gravitational potential. This results in a very small deviation from perfect 
conservation. It is not usually larger than the other effects which result in energy non-conservation 
for our simulations, such as resetting the state of zones that acquire a negative internal energy, and 
in principle if desired it can be made smaller by using stricter tolerance levels on the gravity solve.

@@

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in the appendix, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers. We use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations. A slight 
difference is that the Coriolis force for each velocity component is coupled to other velocity 
components. If the rotation is about the $z$-axis, then the discrete update to 
$u^{n+1}$ depends on the value of $v^{n+1}$, and vice versa. If we fix the value of 
the time-level $n+1$ quantities after coming out of the hydrodynamics update, there 
would be a slight inconsistency between the $x$ and $y$ components of the velocity.

@@

\section{CASTRO hydrodynamics changes}

@@

>>>>

We use the unsplit piecewise-parabolic method (PPM) solver in \castro\
to advance the hydrodynamics system in time \citep{ppmunsplit}.  A
number of changes were made to the solver, which are detailed in \autoref{app:hydro}.
These changes bring the algorithm more in line with that of
\cite{ppm}. \castro\ as originally released featured a slightly modified
version of the higher resolution limiters of
\cite{colella_sekora:2008}, which can be accessed in the code using
\texttt{castro.ppm\_type = 2}. The advantage of this limiter is that
it preserves physical extrema rather than clipping them off as in the
original approach of \cite{ppm}. Despite the advantages of this limiter 
we have found it to be unsatisfactory for our problem. There are many regions in our
problem with large density gradients (such as the interface between
the star's atmosphere and the ambient gas outside of it) and in these
regions the algorithm can yield negative densities. This often results
from the limiters interpreting these gradients as being true
minima. As a result, we use the original limiter, which is strictly
monotonicity preserving in the parabolic profiles it generates; this
is activated with \texttt{castro.ppm\_type = 1}.

@@

There are two ways that these source terms enter the system evolution. 
First, during the hydrodynamics update, we alter the edge states that enter
into the determination of the fluxes. (This only applies for the momentum source term;
the gravitational force does not directly do work on the internal energy, which is used 
to infer the pressure.) To second order in space and time, 
this can be done using the cell-centered
source term evaluated at time-level $n$. We choose a more accurate approach, 
which is also second order, of tracing
under the source term; the details of this are described in 
\autoref{app:hydro}. Second, after the hydrodynamics step, we add the time-centered source terms
to the state. First we will describe how we do this for the momentum,
and then we will describe our approach for the energy. This discussion will be somewhat detailed.
We believe that the attention is necessary because of the importance of accuracy
in the gravitational source terms for our problem. The stability of the white dwarf binary
system is dependent in large part upon accurate coupling of the hydrodynamics and gravity;
an error in this approach could lead to, for example, a spurious mass transfer episode
that might lead us to very different conclusions about the long term stability of such a system.
Such considerations are generally unimportant for spherically-symmetric single star calculations,
but are of the utmost importance in a simulation where the global gravitational field can change 
quite significantly over the course of the simulation.

@@

Finally, then, we write the update in a zone for the total energy that conserves $(\rho E_{\text{tot}})$ as:
\begin{equation}
  \Delta (\rho E) = -\frac{1}{2}\sum_{f} \Delta \rho_{f} (\Phi_{f+1/2} - \Phi_{f-1/2}),\label{eq:grav_energy_conservation_update}
\end{equation}
where the sum is over the cell faces with indices $f$ and the indices $f+1/2$ and $f-1/2$ refer to 
the zone centers immediately to the left and right in the direction perpendicular to the face.
As long as we record the hydrodynamical fluxes through the zone faces after coming out of the hydrodynamics step, 
this algorithm is able to conserve the total energy completely (except for any energy loss or gain through 
physical domain boundaries). In order for the method to be second-order accurate in time, 
we need to use a time-centered $\Phi$ (which can be computed by averaging the time-level $n$ and $n+1$ potentials;
we already have the latter because \castro\ re-computes the potential at the new time after the hydrodynamics step,
and we can apply this energy at the end of the timestep). Note that of course the hydrodynamical
flux is already second-order accurate in time. We observe also that in practice we will not obtain 
conservation of energy to machine precision even in the absence of open domain boundaries. The 
method itself is conservative if it is time-centered and correctly evaluates the energy change 
on cell faces. This was demonstrated empirically by \cite{jiang:2013} and is obvious in the case of a
fixed external potential; it is not as obvious in the case of the gravitational potential, which
changes in response to the mass distribution, so we give a short proof of this in \autoref{app:gravity}.
However, in practice there is a non-zero numerical tolerance associated 
with the Poisson gravity solver (in our case, the multigrid method) that results in a non-zero error 
in the calculation of the gravitational potential. This results in a very small deviation from perfect 
conservation. It is not usually larger than the other effects which result in energy non-conservation 
for our simulations, such as resetting the state of zones that acquire a negative internal energy, and 
in principle if desired it can be made smaller by using stricter tolerance levels on the gravity solve.

@@

In a rotating reference frame with angular frequency vector $\bm{\omega}$, the momentum equation becomes:
\begin{equation}
  \left.\frac{\partial(\rho \mathbf{u})}{\partial t}\right|_{\text{rot}} = -2\, {\bm\omega} \times (\rho\mathbf{u}) - \rho {\bm\omega} \times \left({\bm\omega} \times \mathbf{r}\right).
\end{equation}
Here $\mathbf{r}$ is the position vector with respect to the origin. Typically we choose $\bm{\omega} = (0, 0, 2\pi / T)^T$,
with the rotation axis coincident with the $z$ axis at $x = y = 0$.
$T$ is the rotation period, which is the most natural quantity to specify
for a rotating stellar system. As described in \autoref{app:hydro}, we include this source term
in the edge state prediction in a way that is analogous to the gravity source.
We evaluate all quantities at cell centers. We use the same predictor-corrector 
approach that we use for the gravity source terms to the momentum equations. A slight 
difference is that the Coriolis force for each velocity component is coupled to other velocity 
components. If the rotation is about the $z$-axis, then the discrete update to 
$u^{n+1}$ depends on the value of $v^{n+1}$, and vice versa. If we fix the value of 
the time-level $n+1$ quantities after coming out of the hydrodynamics update, there 
would be a slight inconsistency between the $x$ and $y$ components of the velocity.

@@

\section{CASTRO hydrodynamics changes}
\label{app:hydro}

@@

\section{Proof of Energy Conservation in Simulations using Self-Gravity}
\label{app:gravity}

In \autoref{sec:gravity_hydro_coupling}, we described our approach to updating the gas energy
in response to motions of fluid through the self-generated gravitational potential using 
\autoref{eq:grav_energy_conservation_update}. While it is straightforward to observe that this approach
should be conservative for an arbitrary fixed external potential $\Phi$, it is not as obvious that this
should be so for a self-generated potential which changes in response to mass motions on the domain. To
see that this still holds for the self-generated gravitational potential $\Phi$, let us start with 
\autoref{eq:grav_energy_conservation_update} in a slightly revised form:
\begin{equation}
  \Delta(\rho E)_i = -\frac{1}{2}\sum_{j} \Delta\rho_{ij}(\Phi_i - \Phi_{j}) \label{eq:grav_energy_conservation_update_revised}
\end{equation}
where by $\Delta \rho_{ij}$ we mean the density transferred from zone $j$ to zone $i$, so that
$\Delta \rho_{ij} = - \Delta \rho_{ji}$, and the sum is over all zone indices $j$ that are adjacent
to zone $i$. Let us define $\Phi_{ij} = \Phi_{ji} = (\Phi_{i} + \Phi_{j}) / 2$ as the potential on the
zone interface between zones $i$ and $j$. Then we have:
\begin{equation}
  \Delta(\rho E)_i = -\sum_{j} \Delta\rho_{ij}(\Phi_i - \Phi_{ij}).
\end{equation}
We can evalute the sum for all of the terms proportional to $\Phi_i$ by observing that the change in
density from time-level $n$ to time-level $n+1$ is the sum of the density fluxes from all adjacent zones.
\begin{equation*}
  \Delta(\rho E)_i = - (\rho_i^{n+1} - \rho_i^{n}) \Phi_i + \sum_{j}\Delta \rho_{ij} \ \Phi_{ij}
\end{equation*}
Now let us sum this over all zones $i$ in the domain, and ignore the domain boundaries, or assume that they are
far enough away from the region of compact support for $\rho$ that $\Phi$ is negligible there. As the second
term on the right-hand side is antisymmetric in $i$ and $j$, it cancels when summing adjacent zones, and we have:
\begin{equation*}
  \sum_{i} (\rho E)_i^{n+1} - \sum_{i} (\rho E)_i^{n} = -\frac{1}{2}\sum_{i} (\Phi_{i}^{n+1} + \Phi_{i}^{n})(\rho_i^{n+1} - \rho_i^{n})
\end{equation*}
Note that, as explained the text, we are using a time-centered $\Phi$ to correspond to the mass fluxes
at time-level $n+1/2$. Finally we re-write this in a form where the difference in total energy between time-levels
$n$ and $n+1$ is on the left-hand side and any sources causing this to be non-zero are on the right-hand side:
\begin{align}
  \sum_{i} \left(\rho E + \frac{1}{2}\rho\Phi\right)_i^{n+1} - \sum_{i} \left(\rho E + \frac{1}{2}\rho\Phi\right)_i^{n} &= \frac{1}{2}\sum_{i} \left(\Phi_i^{n+1}\rho_i^{n} - \Phi_i^{n}\rho_i^{n+1}\right) \notag \\
       &= \frac{1}{8\pi G} \sum_{i}\left(\Phi_{i}^{n+1}\nabla^2 \Phi_{i}^{n} - \Phi_i^{n}\nabla^2 \Phi_{i}^{n+1}\right) \label{eq:total_energy_difference}
\end{align}
\autoref{eq:total_energy_difference} expresses total energy conversation if and only if the right-hand side vanishes.
We observe that the right-hand side has the form of a variant of the divergence theorem often called Green's second identity:
\begin{equation}
  \int (\Phi^{n}\nabla^2 \Phi^{n+1} - \Phi^{n+1}\nabla^2 \Phi^{n}) dV = \int \left(\Phi^{n} \nabla \Phi^{n+1} - \Phi^{n+1} \nabla \Phi^{n}\right) \cdot d\mathbf{S}, \label{eq:green_second_identity}
\end{equation}
where $d\mathbf{S}$ is the area element with vector component parallel to the outward normal. The analogous result holds for the
discretized form in \autoref{eq:total_energy_difference}. With the assumptions used above, the right-hand side of
\autoref{eq:green_second_identity} will vanish as the surface integral is evaluated at infinity, where the potential
tends to zero. This concludes the proof that the method is conservative when the potential used at the zone interfaces
is time-centered, even in light of the change of the potential over the timestep due to the mass motion that is causing the change in the energy.

From the above discussion it is straightforward to see exactly why the method will not be fully conservative to machine
precision in practice. First, we cannot simulate the domain out to infinity, so Green's second identity does not hold exactly
and there will be some loss or addition of energy at domain boundaries. Second, \autoref{eq:total_energy_difference} holds in the
continuum limit by using the Poisson equation, but in practice it is not exactly true that $\rho_i = 4\pi G \nabla^2 \Phi_{i}$ due
to small errors in the potential at the level of the tolerances used in the Poisson solver.
